# Title: Survival Analysis Tutorial using R

## Introduction
- Explain the concept and significance of survival analysis in research.
- Highlight the applications of survival analysis in various fields.

## Prerequisites
- List the required R packages for survival analysis (e.g., `survival`, `survminer`, `ggplot2`).
- Provide installation instructions for the packages if needed.

## Dataset Description
- Introduce the dataset used for the tutorial (e.g., clinical trial data, cohort study data).
- Describe the variables included in the dataset, including survival time, event status, and covariates.

## Data Preprocessing
- Load the dataset into R.
- Handle missing values in the dataset (e.g., imputation or exclusion).
- Transform variables if necessary (e.g., log transformation, categorical variable encoding).
- Check for data quality and consistency.

## Kaplan-Meier Analysis
1. Define the survival time and event status variables.




```{r}
source("libraries.R")

# Generate a synthetic survival dataset
set.seed(123)

# Number of observations
n <- 100

# Generate survival times (months)
time <- rexp(n, rate = 0.2) + 5

# Generate event status (0 = censored, 1 = event occurred)
event <- rep(1, n)
event[sample(1:n, n * 0.2)] <- 0

# Generate covariates
covariate1 <- rnorm(n)
covariate2 <- sample(0:1, n, replace = TRUE)
covariate3 <- rpois(n, lambda = 2)

# Create the dataset
dataset <- data.frame(time = time, event = event, covariate1 = covariate1, covariate2 = covariate2, covariate3 = covariate3)

# Print the first few rows of the dataset
head(dataset)

```


```{r}

# Define survival time and event status variables
time <- dataset$survival_time
event <- dataset$event_status

```

2. Compute and plot Kaplan-Meier survival curves for the overall dataset.

```{r}

# Compute Kaplan-Meier survival curves
km_fit <- survfit(Surv(time, event) ~ 1)

# Plot the overall survival curve
ggsurvplot(km_fit, data = dataset, title = "Overall Survival Curve")

```

3. Stratify the survival curves by relevant covariates.

```{r}
# Stratify survival curves by a categorical covariate
km_fit_stratified <- survfit(Surv(time, event) ~ covariate, data = dataset)

# Plot stratified survival curves
ggsurvplot(km_fit_stratified, data = dataset, title = "Stratified Survival Curves")

```

4. Interpret the survival curves and discuss any observed differences.


## Log-Rank Test

1. Perform a log-rank test to compare survival curves between different groups or categories.

```{r}

# Perform a log-rank test
logrank_test <- survdiff(Surv(time, event) ~ group, data = dataset)

# Print the test results
summary(logrank_test)


```

2. Assess the p-value and interpret its significance.
3. Visualize the results using a log-rank test plot.


```{r}

# Create a log-rank test plot
ggsurvplot(logrank_test, data = dataset, title = "Log-Rank Test Plot")

```

## Cox Proportional Hazards Model

1. Fit a Cox proportional hazards model to examine the relationship between covariates and survival time.

```{r}

# Fit a Cox proportional hazards model
cox_model <- coxph(Surv(time, event) ~ covariate1 + covariate2, data = dataset)

# Print the model summary
summary(cox_model)

```

2. Assess the proportional hazards assumption.
3. Interpret the estimated hazard ratios and their significance.
4. Visualize the results using forest plots or coefficient plots.
